---
title: "Assignment 2: Coding Basics"
author: "Alicia Zhao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
#Generating a sequence from 1 to 100 by 4
seq_hundred=seq(1,100,4); seq_hundred

#2. 
#Generating the mean of seq_hundred
mean(seq_hundred)
#Generating the median of seq_hundred
median(seq_hundred)

#3. 
#Seeing if the mean is greater than the median of seq_hundred
mean(seq_hundred)>median(seq_hundred)

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#creating vectors for name of students, test scores, and whether or not they pass the test
student_name <- c("Stacy", "Tom", "Jillian", "Roger") #character vector
test_score <- c(77, 85, 43, 90) #numeric vector
test_pass <- c(TRUE, TRUE, FALSE, TRUE) #logical vector 

#creating a data frame containing the vectors above
testscores_df <- data.frame (student_name, test_score, test_pass)
names(testscores_df) <- c("Student", "Score", "Pass"); testscores_df
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix is a two-dimensional structure that contains elements of the same type, whereas a data frame is more general and can have columns with different modes. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#creating function to determine whether a test score is passing
passtest <- function(x){
    ifelse(x>=50, TRUE, FALSE)
}

passtest2 <- function(x) {
  if(x>=50) {
    TRUE
  } 
  else {
    FALSE
  }
}

#appying function to test_scores
passtest(test_score)
passtest2(test_score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked. 'if' and 'else' did not work because the function requires a length-one logical vector. However, the vector has length >1 and thus only the first element will be used. 

