---
title: "Assignment 8: Time Series Analysis"
author: "Alicia Zhao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW
This exercise accompanies the lessons in Environmental Data Analytics on time series analysis. 

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme
* Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Call these GaringerOzone201*, with the star filled in with the appropriate year in each of ten cases.

```{r, message = FALSE}
getwd()
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

GaringerOzone2010 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv")
GaringerOzone2011 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv")
GaringerOzone2012 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv")
GaringerOzone2013 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv")
GaringerOzone2014 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv")
GaringerOzone2015 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv")
GaringerOzone2016 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv")
GaringerOzone2017 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv")
GaringerOzone2018 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv")
GaringerOzone2019 <- read.csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv")

```

## Wrangle

2. Combine your ten datasets into one dataset called GaringerOzone. Think about whether you should use a join or a row bind. 

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-13 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. 

```{r}
# 2 

GaringerOzone <- rbind (GaringerOzone2010, GaringerOzone2011, GaringerOzone2012, 
                        GaringerOzone2013, GaringerOzone2014, GaringerOzone2015,
                        GaringerOzone2016, GaringerOzone2017, GaringerOzone2018,
                        GaringerOzone2019)

# 3
GaringerOzone$Date <- as.Date(GaringerOzone$Date, format = "%m/%d/%Y")

# 4
GaringerOzone.filtered <-select(GaringerOzone, Date, Daily.Max.8.hour.Ozone.Concentration,
                                DAILY_AQI_VALUE)


# 5
Days <- as.data.frame(seq(as.Date("2010-01-01"), as.Date("2019-12-31"), "days"))
names (Days) [1] <- "Date"

# 6
GaringerOzone.Date <- left_join(Days, GaringerOzone.filtered)


```

## Visualize

7. Create a ggplot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly.

```{r}
GaringerOzone.plot <- ggplot(GaringerOzone.Date, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) + 
  geom_line() + 
  labs (x = "", y = "Ozone Concentration (ppm)") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
print (GaringerOzone.plot)

```

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration.

9. Create a new data frame called GaringerOzone.monthly that contains aggregated data: mean ozone concentrations for each month. 

10. Generate a time series called GaringerOzone.monthly.ts, with a monthly frequency that specifies the correct start and end dates. 

11. Run a time series analysis. 

12. To figure out the slope of the trend, run the function `sea.sens.slope` on the time series dataset. 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. No need to add a line for the seasonal Sen's slope; this is difficult to apply to a graph with time as the x axis. Edit your axis labels accordingly.

```{r}
# 8
GaringerOzone.Date$Ozone <- na.approx(GaringerOzone.Date$Daily.Max.8.hour.Ozone.Concentration)

# 9
GaringerOzone.monthly <- GaringerOzone.Date %>%
    mutate(Year = year(Date), 
         Month = month(Date)) %>%
  drop_na(Daily.Max.8.hour.Ozone.Concentration) %>%
  group_by (Year, Month) %>%
  summarise (Ozone.Concentration = mean(Daily.Max.8.hour.Ozone.Concentration))

GaringerOzone.monthly$Date <- as.Date(paste(GaringerOzone.monthly$Year, 
                                                GaringerOzone.monthly$Month, 
                                                1, sep="-"), 
                                          format = "%Y-%m-%d")

# 10
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$Ozone.Concentration, 
                               start = c(2010, 1, 1),
                               end = c(2019, 12, 1), frequency =12)

# 11
GaringerOzone_Decomposed <- stl(GaringerOzone.monthly.ts, s.window = "periodic")
plot(GaringerOzone_Decomposed)

GaringerOzone.trend <- smk.test(GaringerOzone.monthly.ts)
GaringerOzone.trend
summary(GaringerOzone.trend)


# 12
sea.sens.slope (GaringerOzone.monthly.ts)

# 13
GaringerOzone.monthly.plot <- 
ggplot(GaringerOzone.monthly, aes (x = Date, y = Ozone.Concentration)) + 
  geom_point() + 
  geom_line() + 
  labs (x = "", y = "Ozone Concentration (ppm)") + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

print(GaringerOzone.monthly.plot)
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: Ozone concentrations from 2010-2019 display a slightly negative trend across individual months (Seasonal Mann-Kendall, z= -2.25, p = 0.025). 
 
